---
title: "LLMRagent in 10 minutes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLMRagent in 10 minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = identical(tolower(Sys.getenv("LLMRAGENT_RUN_VIGNETTES", "false")), "true")
)
```

An agent is three things: a model (an `LLMR::llm_config()`), a persona (a
system prompt), and machinery around them: memory, tools, and budgets. This
vignette builds one of each component. Examples use the open-weight `gpt-oss-20b` on
Groq; set `GROQ_API_KEY` and `LLMRAGENT_RUN_VIGNETTES=true` to run them.

```{r setup}
library(LLMRagent)
cfg <- LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.7)
```

## A first agent

```{r first}
ada <- agent(
  "Ada", cfg,
  persona = "You are Ada, a meticulous statistician. Answer in one or two sentences."
)
ada$chat("What is overfitting?")
ada$chat("How would I detect it in practice?")   # remembers the thread
ada$usage()
```

`chat()` is stateful: the agent keeps its own memory (last 40 messages by
default; see `?memory` for summarizing and retrieval policies). `reply()` is
the stateless sibling, used internally by conversations. For long answers,
`chat(stream = TRUE)` prints tokens as they are generated:

```{r stream}
ada$chat("Explain cross-validation to a newcomer, in one paragraph.",
         stream = TRUE)
```

## Tools with declared powers

`agent_tool()` exposes an R function and records what kind of effect it may
have. The agent decides when to call it; LLMRagent executes the call and feeds
the result back. This lookup declares read access and a five-call ceiling.

```{r tools}
lookup_gdp <- agent_tool(
  function(country) {
    gdp <- c(chile = 335, uruguay = 81, bolivia = 47)  # USD bn, illustrative
    val <- gdp[tolower(country)]
    if (is.na(val)) "unknown" else paste0("$", val, " billion")
  },
  name = "lookup_gdp",
  description = "Look up a country's GDP in USD billions.",
  parameters = list(country = list(type = "string", description = "Country name")),
  required = "country",
  side_effects = "read",
  max_calls = 5L
)

analyst <- agent("Analyst", cfg, tools = lookup_gdp,
                 persona = "A careful economic analyst. Use tools for any figure.")
analyst$chat("Compare the GDPs of Chile and Uruguay using the lookup tool.")
analyst$trace()   # every model call and tool call, with tokens and timing
```

## Budgets: call ceilings and recorded-use gates

The call and tool counters are checked before each actual model round or tool
execution. Token and elapsed-time limits stop the next round after recorded use
reaches the gate. A refusal raises a typed error.

```{r budgets}
frugal <- agent("Frugal", cfg, budget = budget(max_calls = 2))
frugal$chat("one")
frugal$chat("two")
tryCatch(frugal$chat("three"),
         llmragent_budget_error = function(e) "stopped by budget, as designed")
```

## Structured answers

```{r structured}
schema <- list(
  type = "object",
  properties = list(stance = list(type = "string",
                                  enum = list("support", "oppose", "unsure")),
                    reason = list(type = "string")),
  required = list("stance", "reason")
)
ada$ask_structured("Should small samples use t or z intervals?", schema)
```

## Agents calling agents

`agent_as_tool()` turns an agent into a tool, so another agent can consult
it. The supervisor decides for itself when to delegate; each consultation
runs on the specialist's own meter (its `usage()`, its `budget()`).

```{r delegation}
stat <- agent("Stat", cfg,
              persona = "A PhD statistician. Precise about assumptions.")
hist <- agent("Hist", cfg,
              persona = "An economic historian. Institutional context.")

lead <- agent("Lead", cfg,
              persona = "A research lead. Consult specialists, then synthesize.",
              tools = list(agent_as_tool(stat), agent_as_tool(hist)))

lead$chat("Crime fell while policing budgets rose, across many cities.
           What would it take to argue causality?")
stat$usage()   # the consultation showed up here
```

## Pipelines: a fixed sequence of specialists

When the routing is fixed rather than model-chosen, chain agents with
`agent_pipeline()`: each stage transforms the previous stage's output, and
every intermediate product is kept.

```{r pipeline}
run <- agent_pipeline(
  list(
    agent("Extractor", cfg, persona =
      "Extract every factual claim as a numbered list. Nothing else."),
    agent("Checker", cfg, persona =
      "Mark each numbered claim VERIFIABLE or VAGUE, one line each."),
    agent("Editor", cfg, persona =
      "Rewrite the original passage keeping only VERIFIABLE claims.")
  ),
  input = "Our app doubled retention, won three design awards, and users love it."
)
run$output
run$steps      # step, agent, input, output -- the complete step record
```

## A two-agent conversation

Conversations run over a shared, speaker-attributed transcript: every agent
sees the full dialogue each turn, and the transcript comes back as a tidy
tibble, ready for text analysis.

```{r conversation}
rosa <- agent("Rosa", cfg, persona = "A pragmatic city planner. Concrete and brief.")
hugo <- agent("Hugo", cfg, persona = "A skeptical economist. Numbers first. Brief.")

conv <- conversation(
  list(rosa, hugo),
  topic = "Should the city pedestrianize its center?",
  max_turns = 4,
  instruction = "At most three sentences per turn."
)
conv$transcript
```

From here: `vignette("designed-conversations")` tours the ready-made study
formats (debates, focus groups, interviews, deliberations);
`vignette("deliberation-experiment")` runs a complete factorial study with
`agent_experiment()`; and `vignette("fanout-synthesis")` shows strong-plus-cheap
model orchestration with `agent_fanout_synthesis()`. For a per-call record of
everything an agent did, turn on `LLMR::llm_log_enable()` before running.
