---
title: "Governed tools and human review: letting an agent act, safely"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Governed tools and human review: letting an agent act, safely}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = identical(tolower(Sys.getenv("LLMRAGENT_RUN_VIGNETTES", "false")), "true")
)
```

A tool lets an agent act beyond generating text. Often that is the point: you
want the agent to query a database, call an API, or run a calculation. But the
same mechanism can delete a row, send a message, or spend money, and an agent
that decides for itself when to take those actions needs explicit limits. This
vignette covers three of them: declaring what a tool may do, checking what flows
in and out, and pausing for a human before an irreversible step.

```{r setup}
library(LLMRagent)
cfg <- LLMR::llm_config("groq", "openai/gpt-oss-20b", temperature = 0.3)
```

## A governed tool declares its powers

`agent_tool()` wraps an ordinary R function and asks you to state, up front, what
kind of thing it is. A read is not a write; a call that reaches the network is
not a call that stays local. Declaring this gives the later checks a policy to
enforce, and it is recorded in the study manifest so the tool's policy is part of
the apparatus you archive.

```{r tool}
lookup_population <- agent_tool(
  function(city) {
    pops <- c(Chicago = 2.7, Detroit = 0.6, Austin = 1.0)  # millions
    pops[[city]] %||% NA_real_
  },
  name = "lookup_population",
  description = "Population of a US city in millions.",
  parameters = list(city = list(type = "string")),
  required = "city",
  side_effects = "read",   # this tool only reads
  max_calls = 5            # and may run at most five times
)
```

The limits are enforced, not advisory. A sixth call is refused and the model is
told why, rather than executed. A tool that runs too long or returns too much can
be capped through `timeout_s` and `max_bytes`; a timed tool requires `callr` at
construction. Drop the tool into an agent exactly as you would any tool:

```{r use-tool}
demographer <- agent("Demographer", cfg, tools = list(lookup_population),
                     persona = "Use the lookup tool for any population figure.")
demographer$chat("Which is larger, Chicago or Austin, and by how much?")
```

## Guardrails check what flows in and out

A guardrail is a check that runs at a boundary: on the user's input, on the
model's output, or before a tool executes. It can block (raise an error and stop),
warn, or simply flag for the record. Every verdict becomes an event in the run,
so even a flag that does not stop anything leaves a trace.

```{r guardrails}
no_pii <- guardrail(
  "no_email_in_output",
  check = function(payload, context) {
    if (grepl("[[:alnum:]._]+@[[:alnum:].]+", payload)) "output contains an email address" else TRUE
  },
  on_fail = "block",
  stage = "output"
)

careful <- agent("Careful", cfg, guardrails = guardrails(no_pii),
                 persona = "You answer questions about contacting the registrar.")

# a normal answer passes; an answer that leaks an address is blocked
tryCatch(
  careful$chat("Give me a fake example email for the registrar."),
  llmragent_guardrail_block = function(e) "blocked: the output guardrail stopped a leak"
)
```

Tool-stage guardrails receive the tool's `name` and `arguments` before the
function runs. A blocking verdict prevents the side effect. The payload also
has a `result` field, which is `NULL` during this pre-execution check.

## Human approval for the irreversible

Some actions should never happen without a person saying yes. Mark such a tool
`requires_approval = TRUE`. When the model tries to call it, the run does not
proceed and does not silently skip the call. It raises a typed condition
carrying a checkpoint: a complete, serializable snapshot of where the agent was.

```{r gate}
send_email <- agent_tool(
  function(to, body) paste0("SENT to ", to, ": ", body),
  name = "send_email",
  description = "Send an email to a recipient.",
  parameters = list(to = list(type = "string"), body = list(type = "string")),
  required = c("to", "body"),
  side_effects = "external",
  requires_approval = TRUE
)

assistant <- agent("Assistant", cfg, tools = list(send_email),
                   persona = "You help draft and send short emails.")

checkpoint <- tryCatch(
  assistant$chat("Email j.doe@example.edu to confirm Tuesday's meeting."),
  llmragent_pending_approval = function(e) e$checkpoint
)
checkpoint    # shows the pending tool and its arguments
```

A reviewer now decides. The checkpoint is an ordinary R object: you can save it
with `saveRDS()`, hand it to a colleague, and resume tomorrow on another
machine. The checkpoint lets a review continue across sessions and machines.
Approve, reject, or edit the arguments, then resume:

```{r approve}
decided <- approve_tool_call(checkpoint, "approve")
result  <- resume_run(decided)
result$text       # final reply
result$agent      # rebuilt agent with continued accounting
result$checkpoint # the decision that was resumed
```

Rejecting feeds the model a denial and lets it carry on without the action.
Editing lets you fix the arguments -- correct a recipient, trim a message --
before the call runs. Either way the decision is recorded as an approval event,
and because the resumed run keeps the original run id, the archive shows one
continuous study with a human checkpoint in the middle, not two disconnected
fragments.

## Putting it together

The three controls combine. A realistic setup declares each tool's powers, guards
the boundaries against leaks and injections, and gates the few genuinely
dangerous tools behind a person. None of it is forced on a simple study: an
agent with no governed tools, no guardrails, and no gates behaves exactly as it
did before. They earn their place once an agent can affect anything beyond the
model call.
